---
title: "Class 15: Pertussis mini-project"
author: "Noel Lim (PID: A17652474)"
format: pdf

---

## Background 

Pertussis, aka whooping cough, is a highly infectious lunch disease caused by the bacteria *B. Pertussis*.

The CDC tracks pertussis cases numbers per year. Lets have a closer look at this data:

[CDC data](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

We will use the **datapasta** R package to "scrape" this data into R.


```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2024L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,23544)
       )
```




```{r}
library(ggplot2)

baseplot <- ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()

baseplot
```

Add some landmark developments as annotation to our plot. We include the first whole-cell (wP) vaccine roll-out in 1946.

Let's add the switch to acelular vaccine (aP) in 1996.

```{r}
baseplot +
  geom_vline(xintercept = 1946, col="blue") +
  geom_vline(xintercept = 1995, col="red") +
  geom_vline(xintercept = 2020, col="gray") +
  geom_vline(xintercept = 2003, col="green")
  
```

We went from ~200,000 cases pre wP vaccine to ~1,000 cases in 1976. The US switched to the aP vaccine in 1995. We start to see a big increase in 2004 to ~26,000 cases. 

There is a ~10 year lag from aP roll out to increasing case numbers. This holds true of other countries like Japan, UK etc.

**Key question**: Why does the aP vaccine induced immunity wane faster than that of the wP vaccine?

## CMI-PB

The CMI-PB (Computational Models of Immunity Pertussis Boost) makes available lots of a data about the immune response to Pertussis booster vaccination.

Critically, it tracks wP and aP individuals over time to seee how their immune response changes.

CMI-PB make all their data freely available via JSON format tables from their database.

Let;s read the first one of these tables

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5/subject",
                     simplifyVector = TRUE)

head(subject)
```

> Q1. How many subjects are there in this data set?

```{r}
nrow(subject)
```

> Q2. How many aP and wP individuals are there?

```{r}
table(subject$infancy_vac)
```

> Q3. How many male/female?

```{r}
table(subject$biological_sex)
```

> Q4. Breakdown by biological_sex and race, e.g. how many black female subjects etc.

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Does this do a good job of representing the US populus?

No it is not representative

Let's get more data from CMI-PB, this time about the specimens collected.

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5/specimen",
                      simplifyVector = TRUE)

head(specimen)
```

Now we can join (merge) these two tables `subject` amd `specimen` to make one new `meta` table with the combined data

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

Now read an "experiment data" table from CMI-PB

```{r}
abdata <- read_json("http://cmi-pb.org/api/v5/plasma_ab_titer",
                    simplifyVector = TRUE)

head(abdata)
```

One more join to do of `meta` and `abdata` to associate all the metadata about the individual and their race, biological sex and infincy vaccination status together with antibody levels...

```{r}
ab <- inner_join(abdata, meta)
head(ab)
```

> Q. How many Ab measurements do we have?

```{r}
nrow(ab)
```

How many isotypes

```{r}
table(ab$isotype)
```

How many antigens?
```{r}
table(ab$antigen)
```

Let's focus on the IgG - one of the main antibody types responsive to bacteria or viral infections

```{r}
igg <- filter(ab, isotype=="IgG")
head(igg)
```

Make a first plot of MFI (Mean Fluorescence Intensity - measure of how much is detected) for each antigen.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

Let's color by aP/wP infancy_vac

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~visit)
```

```{r}
table(igg$visit)
```

Looks like we don't have data yet for all subjects in terms of visits 8 onward. So lets exclude these.

```{r}
igg_7 <- filter(igg, visit %in% 1:7)
table(igg_7$visit)
```

```{r}
ggplot(igg_7) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit, ncol=2)
```

Let's try a different plot. First focus on one antigen, start with PT(Pertussis toxin) and plot visit. or time on the x-axis and MFI_normalized on the y-axis.

```{r}
ggplot(igg_7) +
  aes(planned_day_relative_to_boost, MFI_normalised) +
  geom_point() +
  geom_line() 
```

```{r}
abdata.21 <- ab %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() 
```

Let's finish here for today. We are beginning to see some interesting differences between aP and wP individuals. There is likely lots of other interesting things to find in this dataset...