---
title: "Class 11: Protein Structure Prediction with AlphaFold"
author: "Noel Lim (PID: A17652474)"
format: gfm
---

> Q10. Which of the packages above is found only on BioConductor and not CRAN? 

MSA Package

> Q11. Which of the above packages is not found on BioConductor or CRAN?:



> Q12. T/F? Functions from the devtools package can be used to install packages from GitHub and BitBucket?

True

```{r}
library("bio3d")
library("devtools")
library("BiocManager")
aa <- get.seq("1ake_A")
aa
```


> Q13. How many amino acids are in this sequence, i.e. how long is this sequence?


## Search PDB database for related sequences
```{r}
# b <- blast.pdb(aa)
```

```{r}
# hits <- plot(b)
```

```{r}
# head(hits$pdb.id)
```

```{r}
hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```
## Download related structures
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```
 
```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
ids <- basename.pdb(pdbs$id)
```

```{r}
#plot(pdbs, labels=ids)
```

```{r}
anno <- pdb.annotate(ids)
unique(anno$source)
anno
```
## Performing PCA
```{r}
pc.xray <- pca(pdbs)
#plot(pc.xray, pc.axes = c(1,2))
```

```{r}
rd <- rmsd(pdbs)
```

```{r}
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)

#plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```
## Visualize first principal component
```{r}
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```
## Plotting results with ggplot2
```{r}
library(ggplot2)
library(ggrepel)

df <- data.frame(PC1=pc.xray$z[,1], 
                 PC2=pc.xray$z[,2], 
                 col=as.factor(grps.rd),
                 ids=ids)

p <- ggplot(df) + 
  aes(PC1, PC2, col=col, label=ids) +
  geom_point(size=2) +
  geom_text_repel(max.overlaps = 20) +
  theme(legend.position = "none")
p
```

