---
title: "Class 10: Halloween Candy Project"
author: "Noel Lim (PID: A17652474)"
format: html
---

Today is Halloween, an ole Irish holiday, let's celebrate by eating candy.

We will explore some data all about Halloween candy from the 538 website.

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)
```

> Q. How many candy entries

```{r}
nrow(candy)
```

> Q2. How many fruit candy types are in the dataset

```{r}
sum(candy$fruity)
```

```{r}
sum(candy$chocolate)
```


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Skittles original","winpercent"]
```

```{r}
candy["Rolo",]$winpercent
```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)=="Haribo Happy Cola") |>
  select(winpercent)
```

> Q. Find fruity candy with a winpercent above 50%

```{r}
candy |>
  filter(winpercent > 50) |>
  filter(fruity==1)
```

```{r}
top.candy <- candy[candy$winpercent > 50,]
top.candy[top.candy$fruity==1,]
```

To get a quick insight into a new dataset some folks like using the skimer package and its `skim()` function.

```{r}
skimr::skim(candy)
```

Looks like the `winpercent` variable or column is measured on a different scale than everything else! I will need to scale my data before doing analysis like PCA etc.

> Q8. Plot a histogram of winpercent values

We can do this a few ways, e.g. the "base" R `hist()` function or with `ggplot()`

```{r}
hist(candy$winpercent, breaks=100)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth = 8) +
  theme_bw()
```


> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
fruit.candy <- candy |>
  filter(fruity==1)

summary(fruit.candy$winpercent)
```

```{r}
#choco.candy <- candy[as.logical(candy$chocolate,]$winpercent

choc.candy <- candy |>
  filter(chocolate==1)
summary(choc.candy$winpercent)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.candy$winpercent, fruit.candy$winpercent)
```

> Q13. What are the five least liked candy types in this set?

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
play <- c("d","a","c")
sort(play)
order(play)

play[ order(play) ]
```

```{r}
sort(c(5,2,10), decreasing = T)
```


```{r}
tail( candy[order( candy$winpercent),],5)
```
Let's do a barplot of winpercent values
```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent),
      fill=chocolate)+
  geom_col()
```

I want a more custom color scheme where I can see both chocolate and bar and fruity etc. all from the one plot. To do this we can roll our own color vector...

```{r}
#Place holder color vector
mycols <- rep("black", nrow(candy))
mycols[as.logical(candy$chocolate)] <- "chocolate"
mycols[as.logical(candy$bar)] <- "brown"
mycols[as.logical(candy$fruity)] <- "pink"
```

```{r}
# Use blue for your favorite candy
mycols[ rownames(candy)=="Twix"] <- "blue"
mycols
```

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```

Plot of winpercent vs pricepercent to see what would be the best candy to buy ...

```{r}
mycols[as.logical(candy$fruity)] <- "red"
```


```{r}
ggplot(candy) +
  aes(winpercent, pricepercent) +
  geom_point(col=mycols)
```

Add labels
```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text(col=mycols)
```

Make the labels non-overlapping
```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, max.overlaps=8)
```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reeses Miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip


```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

# PCA

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,1],pca$x[,2], col=mycols, pch=16)
```

How do the original variables (columns) attribute to the new PCs. I will look at PC1 here

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) +
  aes(PC1, reorder(rownames(loadings), PC1), fill=PC1) +
  geom_col()
```


Let's make a nicer score plot with ggplot. Again I need a data.frame with all the stuff I want (PC results and candy data) for my plot as input

```{r}
pc.results <- cbind(candy,pca$x)

ggplot(pc.results) +
  aes(PC1, PC2, label=rownames(pc.results)) +
  geom_point(col=mycols) +
  geom_text_repel(col=mycols, max.overlaps = 8) +
  labs(title="Candy Space via PCA")
```

